An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-14 paper.tex
Edits to Bayesian Analysis section; formatting.
Changed \begin{equation} to \be, etc.  Fixed some language (model
evidence --> model posterior probability, etc).  Added Weinberg to
references.

@@

Consider an observed data set $d$ and a set of competing models for the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.  Each model has some continuous parameters, $\vtheta_i$; given the model and its parameters, we can make a prediction about the likelihood of observing
the experimental data: $L(d|\vtheta_i, M_i)$.  Within the framework of each model, Bayes' rule gives us a way to compute the posterior probability distribution function (PDF) for the model
parameters implied by the data:
\begin{equation}
  p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
\end{equation}
where $p(\vtheta_i |d, M_i)$ is the posterior distribution for the
model parameters $\vtheta_i$ implied by the data in the context of
model $M_i$, $p(\vtheta_i|M_i)$ is the prior
probability of the model parameters that 
represents our beliefs before accumulating any of the data $d$, and
$p(d|M_i)$, called the evidence, is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
the $\vtheta_i$.  This implies that the evidence is equal to
\bel{evidence}
  p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
\end{equation}
where $V_i$ is the parameter space volume in model $M_i$.  We are interested in the evidence for model $M_i$ given the data, $p(M_i|d)$, which is again given by Bayes' rule as
\begin{equation}
p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
\end{equation}
where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
\be
p(d)=\sum_i p(d|M_i) p(M_i).
\ee

When selecting among alternative models, we are interested in finding the model with the highest evidence $p(M_i|d)$.  However, attempts to directly compute the evidence by performing the integration in Eq.~(\ref{evidence}) are generally very difficult in a multi-dimensional, multi-modal parameter space when the likelihood has to be evaluated numerically.  In particular, a grid-based integral quickly becomes computationally unfeasible as the dimensionality of $\vtheta$ exceeds a few.  The parameter space must typically be explored in a stochastic manner before the evidence integral can be computed.  Although several stochastic parameter-exploration techniques are focused directly on evidence computation (e.g., nested sampling and MultiNest \ilya {Add references}), we frequently want to compute the posterior PDFs within each model along with the evidences for the various models, and the Markov chain Monte Carlo is one of the most widely used tools for this purpose.

@@

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\be
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1})=\vtheta^{(i)}$.
 \end{enumerate}

>>>>

\will{Mention Weinberg direct integration (Ref.~\cite{Weinberg2009}).}

@@

Consider an observed data set $d$ and a set of competing models for
the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.
Each model has some continuous parameters, $\vtheta_i$; given the
model and its parameters, we can make a prediction about the
likelihood of observing the experimental data: $L(d|\vtheta_i, M_i)$.
Within the framework of each model, Bayes' rule gives us a way to
compute the posterior probability distribution function (PDF) for the
model parameters implied by the data:
\be
  p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
\ee
where $p(\vtheta_i |d, M_i)$ is the posterior distribution for the
model parameters $\vtheta_i$ implied by the data in the context of
model $M_i$, $p(\vtheta_i|M_i)$ is the prior probability of the model
parameters that represents our beliefs before accumulating any of the
data $d$, and $p(d|M_i)$, called the evidence, is an overall
normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is
properly normalized as a probability distribution on the $\vtheta_i$.
This implies that the evidence is equal to
\bel{evidence}
  p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
\ee
where $V_i$ is the parameter space volume in model $M_i$.  For model
comparison, we are interested in the posterior probability of a
particular model, $M_i$, given the data, $p(M_i|d)$.  Using Bayes'
rule, we see that this involves the evidence (Eq.~\eqref{evidence}):
\be
p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
\ee
where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a
normalizing constant,
\be
p(d)=\sum_i p(d|M_i) p(M_i).
\ee

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling and MultiNest \ilya {Add references}).  However, we
frequently want to compute the posterior PDFs within each model along
with the evidences for the various models.  One of the most common
techniques for computing posterior PDFs in the context of a model is
the Markov Chain Monte Carlo, which we now describe.

@@

One way to produce such a sequence of samples is via the Metropolis-Hastings algorithm, first proposed by Metropolis et al.~in 1953 \cite{Metropolis:1953}, and later generalized by Hastings:
\begin{enumerate}
  \item Given a current state $\vtheta^{(i)}$, propose the next state $\vtheta^p$ by drawing from a jump proposal distribution with probability $Q(\vtheta^{(i)} \to \vtheta^p)$.  
  \item Compute the probability of accepting the proposed jump as
\be
p_{\textnormal{accept}}  \equiv \min\Bigl(1,  \frac{p(\vtheta^p|d)}{p(\vtheta^{(i)}|d)} \frac{Q(\vtheta^p \to
        \vtheta^{(i)})}{Q(\vtheta^{(i)} \to \vtheta^p)} \Bigr).
\ee
\item Pick a uniform random number $\alpha \in [0,1]$.  If $\alpha<  p_{\textnormal{accept}}$, accept the proposed jump, setting $\vtheta^{(i+1)}=\vtheta^p$.  Otherwise, reject the jump, and remain at the same location in parameter space for the next step, $\vtheta^{(i+1)}=\vtheta^{(i)}$.
 \end{enumerate}
